---
title: 'Cabrillo National Monument: Seal and Sea Lion Park Management Briefing'
author: "L. Pandori"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r packages and themes}

##### load packages #####
library(kableExtra)# more nice tables
library(calecopal) # color palette (sierra 2)
library(lubridate) # dates/times
library(readxl)    # read excel files
library(tidyverse) # dplyr, forcats, etc.

###### set themes #####
ll_theme <- theme_bw() + theme(text = element_text(size = 12, color = 'black'),
                                    # add more space between panels
                                    panel.spacing = unit(1, 'lines'),
                                    # no background to wrap panels
                                    strip.background = element_blank(),
                                    strip.text = element_text(size = 12, 
                                                              hjust = 0),
                                    # panel labels outside x axis labels
                                    strip.placement = 'outside',
                                    # adjust x axis labels
                                    axis.text.y = element_text(size = 12),
                                    axis.text.x = element_text(size = 12, 
                                                               angle = 45, 
                                                               hjust = 1),
                                    panel.grid = element_blank())

# read in pinniped data
pin <- read_csv('../data/pinniped_summary.csv')

# get complete data without duplicates
pin <- pin %>%
          na.omit() %>%
          distinct()

```

### Introduction

Cabrillo National Monument (CABR) is a unit of the National Park System located at the tip of Point Loma in San Diego, CA. Despite its small footprint (160 terrestrial acres), the monument attracts nearly one million visitors each year, 380,000 of which visited the coastal area of the park in 2020. Further, the *c.* 1 mile stretch of shoreline in the coastal area of the park is used by seals and sea lions. This results in potential conflicts between conservation of natural resources, visitors, and other users of the area. Due to its value as both a habitat refuge and public resource, it is critical to monitor use patterns by visitors, seals and sea lions at Cabrillo National Monument.

### Objectives

This program provides estimates of seal, sea lion, and visitor use at two sites in the coastal area of Cabrillo National Monument: the vista overlooking rocky intertidal Zone 1 (closest to Lot 1 and the tidepool area), and the Spur Trail Vista overlooking the Sea Cove area. Surveys were conducted in June 2021 through January 2022. Specifically, this study: (1) documents trends in seal, sea lion and visitor use over time, and (2) identifies preferred haul-out sites (places where seals and sea lions rest on land).

### Results

The average number of visitors per survey is higher at the Zone 1 Vista than at the Spur Trail Vista. Alternately, the average number of seals is higher at the Spur Trail Vista than the Zone 1 Vista. Only `r sum(pin$int_lion_adult) + sum(pin$swim_lion)` sea lions were observed during `r length(unique(pin$date))`` surveys conducted between June 2021 and January 2022. Since there were so few sightings, the following charts show only seal abundance over time and seal haul-out (land rest) site preference. 

```{r totals of observations}

# wrangle data for table - calculate observation totals for study period
obs_table <- pin %>%
  select(vista:humans) %>%
  # pivot longer to do calculations on rows
  pivot_longer(., cols = int_seal_adult:humans, names_to = 'obs_type') %>%
  mutate(obs_type = case_when(grepl('lion', obs_type) ~ 'Sea lions',
                              grepl('seal', obs_type) ~ 'Seals',
                              TRUE ~ 'Visitors')) %>%
  group_by(vista, obs_type) %>%
  # get sum of each type normalized by survey dates
  summarize(n = round(sum(value)/length(unique(pin$date)), digits = 0)) %>%
  # pivot wider into a table with vista along the columns and type in rows
  pivot_wider(names_from = vista, values_from = n) %>%
  ungroup() %>%
  # nice names
  rename(`Zone 1` = zone1, `Spur Trail` = spur, `Lot 3` = lot3, `Observation type` = obs_type) %>%
  # reorder from S to N
  select(`Observation type` ,`Zone 1`, `Spur Trail`, `Lot 3`) %>%
  # make obs type column blank for easier to read table
  rename(` ` = `Observation type`) %>%
  # remove lot 3 vista
  select(-`Lot 3`)

# make kable table output
kable(obs_table, 
      # formatting
      format = 'html', align = 'c', 
      # caption with number of surveys
      caption = 'Average count per survey') %>%
  # add more styling
  kable_styling(# differentiate rows w color and cols with borders
                bootstrap_options = c('bordered', 'striped'),
                # doesn't need to take up full pg width
                full_width = F,
                # left-align in document
                position = 'left',
                font_size = 12) %>%
  # top and bottom borders
  row_spec(0, extra_css = "border-top: 1px solid #000000; border-bottom: 1px solid #000000;") %>% 
  row_spec(nrow(obs_table), extra_css = 'border-bottom: 1px solid #000000;')

remove(obs_table)

```

While most seals observed from the Zone 1 Vista were swimming in the ocean, while those in observed in the Sea Cove from the Spur Trail Vista were hauled-out. Only `r sum(pin$int_seal_pup) + sum(pin$int_lion_pup)` pups were observed between `r paste(month(min(pin$date), label = T, abbr = F), year(min(pin$date)))` and `r paste(month(max(pin$date), label = T, abbr = F), year(max(pin$date)))``. 

```{r abundance over time}

pin_plot <- pin %>%
  # nice vista names (leave lot 3 in case we include it in the future)
  mutate(vista = case_when(vista == 'lot3' ~ 'Lot 3',
                           vista == 'spur' ~ 'Spur Trail',
                           vista == 'zone1' ~ 'Zone 1')) %>%
  # select cols of interest
  select(date, vista, int_seal_adult:humans) %>%
  # pivot longer so we can filter obs types using grepl
  pivot_longer(int_seal_adult:humans, names_to = 'obs_type', values_to = 'n') %>%
  # get seal values only + filter for not lot 3
  filter(vista != 'Lot 3' & grepl('seal',obs_type)) %>%
  # add moth column
  mutate(mo = floor_date(date, unit = 'month')) %>%
  # get avg obs/survey for each month
  group_by(mo, vista, obs_type) %>%
  summarize(n = mean(n)/length(n)) %>%
  # PLOT STARTS HERE
  ggplot(data = .,
       mapping = aes(x = mo, y = n, group = fct_rev(obs_type), fill = fct_rev(obs_type))) + 
  geom_col(color = 'black') + 
  scale_fill_manual('Seal activities',
                     values = cal_palette('halfdome', n = 3, type = 'discrete'), 
                     # add custom labels to values
                     label = c('Swimming', 'Hauled-out pup', 'Hauled-out adult')) +
 #scale_x_date(labels = format('%m')) +
  xlab('Month') + 
  ylab('Average observations per survey') + 
  facet_wrap(~fct_rev(vista)) +
  ll_theme

plot(pin_plot)
  
```

### Methods Overview

California sea lions (*Zalophus californianus*) and harbor seals (*Phoca vitulina*) are the focal species of this sampling due to their high abundances and frequent haul-out, or land resting, behavior along southern California shores (Antonelis & Fiscus 1980). Volunteers in Parks (VIPs) conduct 10 minute observations of the numbers of seals and sea lions visible from three overlooks along the coastal area of the park (Zone 1, Spur Trail and Lot 3 overlooks). After conducting seal and sea lion observations, volunteers conduct 10 minute observations of the number of humans in the intertidal area. They record the time, estimated percent cloud cover, and subjective wind level (on a scale of 0-3) during surveys. The time of survey events is used to get the tide level, air temperature and water temperature from the National Oceanographic and Atmospheric Administration (NOAA) station at Scripps Pier in La Jolla.